Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2013-09-18 Dail_Mad_ext.tex
Edits from Scott plus appendices updated

\author{Jeffrey A. Hostetler\\ %\footnote{Contact:
%     \url{HostetlerJ@si.edu}, +001-202-633-4205}\\
%        \small Migratory Bird Center \\
        \small Smithsonian Conservation Biology Institute \\
        \small National Zoological Park, MRC 5503 \\
        \small Washington, DC 20013-7012\\
   \and Richard B. Chandler \footnote{Contact:
     \url{RChandler@usgs.gov}, +001-301-497-5856} \\
        \small USGS Patuxent Wildlife Research Center \\
        \small Laurel, MD 20708}
\maketitle

\abstract{
\noindent
\begin{enumerate}
\item Models of population dynamics play a central role in theoretical and
applied ecology, where they are used for purposes such as testing 
hypotheses about density dependence and predicting species' responses 
to future environmental change or conservation actions. Failure to
account for observation error in such models can result in 
bias, and thus ecologists have increasingly relied on state-space
models to directly model both the observation error and the ecological
process of interest.
\item Conventional state-space models, however, have
three important limitations: (1) the parameters are not identifiable
in many common situations, (2) they do not admit spatial variation in
population dynamics, and (3) there is no clear interpretation of the
observation error. We demonstrate how each of these problems can be
resolved using a class of hierarchical models proposed by
\citet[Biometrics]{dail_madsen:2011} for spatially-replicated
time-series data that attributes observation error to imperfect detection.
\item We expand this class of models to accommodate classical growth
models (e.g. exponential and Ricker-logistic), zero-inflation,
and random effects such as observer-specific
detection probabilities. We also present methods for forecasting
population size under future environmental conditions. Implementation
of these ideas is possible using either frequentist or Bayesian
methods, and code to fit these models using the \textbf{R} package
\texttt{unmarked} and \textbf{JAGS} is also provided.
\item A simulation study indicated that bias was negligible and coverage
nominal for the proposed model extensions. An analysis of data from
the North American Breeding Bird Survey highlighted how these methods
can be readily applied to existing data, but it also suggested the
that precision will be low when direct information about detection
probability (such as is collected using distance sampling or
replicated counts) is unavailable.
\end{enumerate}
}

@@

Theoretical ecology requires models of population dynamics for testing
hypotheses regarding spatial and temporal variation in
abundance. For example, much theoretical work has focused on
understanding the importance and existence of phenomenon such as
density-dependent population regulation, population cycling, and spatial synchrony
\citep{may:1975,royama:1977,turchin:1990,dennis_taper:1994,bjornstad_etal:1999},
and population models are required to evaluate associated hypotheses.
In applied contexts, population models are used for estimating extinction probabilities
\citep{schoener_spiller:1992,nadeem_lele:2011,hostetler_etal:2012} and for predicting the
effects of future environmental conditions or conservation actions on
population size \citep{jamieson_brooks:2004,hatfield_etal:2012}.
In order to address these questions, two complicating factors must
be confronted when fitting population models to data. 
First, deterministic models of population dynamics are virtually always inadequate due to process
variation, the inherent stochasticity in demographic parameters and environmental
conditions \citep{bjornstad_grenfell:2001,saether_engen:2002}.
Second, abundance---the natural state variable in studies
of population dynamics---can rarely be observed perfectly in field
studies because of observation error, such as imperfect
detection \citep{link_nichols:1994,kery_etal:2009}.

@@

A simple state-space models can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a population growth model, $\mu(N_{t-1})$
and random process variation denoted $\grave{o}_t$.
The growth model may be density-dependent, as in the case of the 
Ricker-logistic model, or it might be density-independent, such as when growth
is exponential.
%, in which $\mu(N_{t-1}) = N_{t-1}e^{r}$ where $r$ is the intrinsic rate of increase.
The full model can now be written as:
\begin{subequations}
  \label{eq:ss1}
  \begin{align}
    N_1 &= X_1 \label{eq:ss1a} \\
N_t &= \mu(N_{t-1}) + \grave{o}_{t-1} \quad \text{for} \;
t=2,\hdots,T \label{eq:ss1b} \\
X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \;
t=1,\hdots,T \label{eq:ss1c}
  \end{align}
\end{subequations}
where $\grave{o}_t$ is the random effect allowing for process
variation unaccounted for by the deterministic model. In classical
state-space models, the two sets of random effects
are assumed to be i.i.d. Gaussian deviates:
$\grave{o}_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. It is also standard practice to
ignore process variation associated with $N_1$, as
indicated by Eq~\ref{eq:ss1a}. Alternatively, the population is
assumed to be at equilibrium such that the $N_1$ can be regarded as an
outcome of the equilibrium distribution.

Even though state-space models such as that shown in Eq.~\ref{eq:ss1}
are among the most widespread approaches for modeling population dynamics
using time-series data, several problems are evident. Namely, (1)
time-series data and the underlying abundance parameters of interest
are typically integer-valued, as in the case of count data, raising
concerns about the use of Gaussian distribution for the random
effects, (2) the model for observation error has little biological basis, (3)
spatial variation in abundance is not allowed, and (4) some of the
parameters of the model are not estimable. We briefly discuss each of
these points before describing a general approach to resolving these
shortcomings.

Use of the Gaussian distribution for modeling random process variation
and observation error is motivated by convenience rather than
biology. Specifically, the Gaussian assumptions
allow for parameter estimation using the Kalman filter
\citep{dennis_etal:2006}, which is much more computationally efficient
than estimation methods when random effects are not Gaussian distributed
\citep{devalpine_hastings:2002}. The problem with this is that it
allows for negative and non-integer values of the state-variable, which is
inconsistent with the observed data, and may result in implausible predictions.
Transforming the count data and the underlying abundance variables do
little to resolve these issues \citep{ohara_kotze:2010}.

Another problem with standard state-space models is that the
observation error has no clear interpretation. The use of a mean zero
Gaussian distribution implies that $X_t$ will be
higher than $N_t$ as often as it is lower than $N_t$. 
It is hard to identify a mechanism that would cause such symmetric errors.
A more likely form of observation error, and one that has been recognized for well
over a century, results from failing to detect individuals that are
present. Imperfect detection may be attributable to
characteristics of the species under study, such as its elusiveness,
or to the failings of the ecologist collecting the data in the field.
Although a vast number of methods have been devised for accounting for
this form of observation error, rarely have these methods been
integrated into state-space models \citep[but
see][]{buckland_etal:2004}.

A more serious problem %than the ones associated with the Gaussian assumptions 
is that the parameters of the simple state-space models
such as Eq~\ref{eq:ss1} are not identifiable in many
circumstances \citep{polansky_etal:2009}. Recall that the
Eq.~\ref{eq:ss1} did not specify distributions for $N_1$; % or $X_1$;
however, $N_1$ is a random variable and hence there is uncertainty
that should be accounted for.  
%are needed to fully specify the model.
%To be more specific, a fully-specified state-space model requires at least
%three probability distributions, which we represent
%using bracket notation (i.e., $[Y|\Omega]$ is interpreted
%as the probability distribution of the random variable $Y$ given the
%parameter $\Omega$). The three probability distributions required for a
%state-space model are:
%%%% JEFF, does this extra sentence help? This notation comes from
%%%% several of the original state-space papers and is widespread in
%%%% books like Andy's and Bill's.
%\hl{[I'm not familiar with bracket notation,
% so these equations are confusing to me. How widely used is this
% notation in ecology? Will others likely understand this? Also, we might want
%to save space by trimming a bunch of this discussion back down.]}
%\begin{subequations}
%  \label{eq:ss2}
%  \begin{align}
%    [N_1&|\bm{\theta}] \label{eq:ss2a} \\ \label{eq:ss2b}
%[N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\
%\label{eq:ss2c}
%    [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
%  \end{align}
%\end{subequations}
%where $\bm{\theta}$ are the process variation parameters for the
%initial state, $\bm{\Theta}$ are the process variation parameters
%influencing how abundance changes over time, and $\bm{p}$ are the
%observation error parameters. As mentioned previously, Eq.~\ref{eq:ss2b}
%and Eq.~\ref{eq:ss2c} are assumed to be Gaussian in classical state-space
%models, but what should be the distribution for Eq.~\ref{eq:ss2a}?
%The three lines of general model
%described by Eq.~\ref{eq:ss2} correspond to the three equations shown
%for the specific example in Eq~\ref{eq:ss1}. However, in this case,
%we allow $N_1$ to be a random variable. But what should its
%istribution, $[N_1|\bm{\theta}]$ be?
%And how could $\bm{\theta}$ be estimated given that there is only a single observation
%available? Often, these problems are simply assumed away. For
%example, sometimes researchers assume that there is no process variation in
%the first year, although this seems unjustified. Alternatively, the
%population may be assumed to be at equilibrium such that the expected
%value of $N_t$ is constant through time. Although this 
%makes the parameters $\bm{\theta}$ identifiable, assuming equilibrium
%defeats the objective of many studies of population dynamics, namely
%determining why a population varies over time.

A fourth problem with these models is that they do not admit
spatial variation. This reduces the scope of the inferences that can
be drawn from the models and it ignores the importance of space in
population regulation.

@@

The DM model is an extension of the $N$-mixture model
\citep{royle:2004biom}, which allows for inference about spatial
variation in abundance when individuals cannot be detected with
certainty. To estimate both abundance parameters and parameters
of the detection process, the original $N$-mixture model uses replicate
observations at each site, which are collected during sufficiently
short time intervals such that the population can safely be
assumed to be closed with respect to births, deaths, and movement. The DM
model relaxes this closure assumption and includes explicit parameters
describing population change over time.

@@

As with traditional state-space models, the DM model includes the
three conditionally related processes shown in Eq.~\ref{eq:ss1},
which correspond to (1) initial abundance, i.e. the
abundance at site $i$ during the first primary period,
denoted (2) abundance at time $t$ (for $t>1$) which depends upon
abundance at $t-1$, and (3) the
detection process \citep{dail_madsen:2011}.
The first two processes describe the state process---the
variation in abundance in space and time. The third process %simply 
describes the the relationship between
abundance and the observed count data.

@@

The DM model assumes that abundance in time $t$ is a function of
abundance in time $t-1$, i.e. abundance at each site evolves as a
first order Markovian process. %, although higher order processes are also possible.
\citet{dail_madsen:2011} considered several models to describe the temporal dynamics;
however, in each case they modeled $N_t$ as the sum of two random variables:
$S_{it}$, the number of indivduals surviving from $t-1$ and not
emigrating; and $G_{it}$ the number of new individuals entering
the population. Their most general model was
\begin{equation}
\left.\begin{aligned}
S_{it}|N_{it-1} &\sim \mathrm{Bin}(N_{it-1}, \omega) \\
G_{it}|N_{it-1} &\sim \mathrm{Pois}(\gamma(N_{it-1})) \\
N_{it} &= S_{it}+G_{it}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
where $\omega$ is the apparent survival probability and $\gamma$
is the recruitment rate (which can depend on $N_{it-1}$).
\citet{dail_madsen:2011} proposed three
models for $\gamma$: the constant model,
$G_{it} \sim \mathrm{Pois}(\gamma)$ where recruitment does not
depend on $N_{it-1}$, and which simulates a ``propagule rain'' of new
individuals; the autoregressive model, $G_{it} \sim
\mathrm{Pois}(\gamma(N_{it-1}))$, which
simulates geometric or density independent growth; and the
``no-trend'' model, $\gamma = (1-\omega)\Lambda$, which keeps
expected abundance constant over time. As before, covariates of
$\omega$ and $\gamma$ can be easily accommodated, for example
using logit- and log-linear models respectively.

@@

However, Bayesian inference has several appealing features. First, it
allows direct probability statements to be made about a hypothesis
given data \citep{link_barker:2010}.    
Second, Bayesian methods offer straight-forward approaches for
combine data from multiple sources or use existing estimates of parameters as prior
distributions. In practice, simulation methods such as Markov
chain Monte Carlo (MCMC) are used to simulate the posterior
distributions. %Although MCMC can be slow, our experience is that it
%may actually converge faster than the amount of time required to
%maximize the likelihood shown in Eq.\ref{eq:like}, especially
%when the upper bound of the summation must be high, e.g. $N_{max}>200$.
%Furthermore, 
MCMC may be the only possible option for estimating parameters
using some of the extensions described below, such as allowing for
additional random effects.

@@

%For example here is \textbf{BUGS} code for the ``geometric'' model described
%in Eq.~\ref{eq:hm}:
%\begin{verbatim}
%model {
%lambda ~ dunif(0, 5)
%omega ~ dunif(0,1)
%gamma ~ dunif(0, 10)
%p ~ dunif(0,1)
%for(i in 1:nSites) {
%  N[i,1] ~ dpois(lambda)
%  y[i,1] ~ dbin(p, N[i,1])
%  for(t in 2:nYears) {
%    S[i,t-1] ~ dbin(omega, N[i,t-1])
%    G[i,t-1] ~ dpois(gamma*N[i,t-1])
%    N[i,t] <- S[i,t-1] + G[i,t-1]
%    n[i,t] ~ dbin(p, N[i,t])
%    }
%  }
%}
%\end{verbatim}

@@

Although partitioning population growth into survival and recruitment
is ideal in terms of providing a mechanistic description of population
dynamics, it is not always possible using simple count data,
especially when the sites are not closed with respect to immigration
and emigration. %Furthermore, a
%simpler model would have several merits: faster running time, fewer
%total model combinations, and possibly more realistic estimates.
If the mechanistic model must be sacrificed, we suggest replacing it
with classical population growth models that have a long history in
ecology. The simplest of which is the exponential growth
model: $f(N_{it-1}) = N_{it-1}e^r$ where $r$ is the intrinsic
rate of increase. %Although demographic and environmental stochasticity
%in $r$ could be modeled as functions of covariates or random effects, a
%simpler way of 
To allow for demographic stochasticity, we can regard $N_{it}$ as a Poisson 
random variable, simplifying Eq.~\ref{eq:Nt} to:
\begin{equation}
  N_{it} \sim \text{Pois}(\exp(r)N_{it-1}).
\label{eq:exp}
\end{equation}
This,
like the geometric-recruitment version of the model, is a variant on a simple
density-independent exponential model of population growth.
Density-dependent versions of the model are also possible.  For
example:
\begin{equation}
  N_{it} \sim \text{Pois}(N_{it-1}\exp(r(1-N_{it-1}/K)))
\label{eq:rick}
\end{equation}
where $K$ is the stable equilibrium of the population and $r$ is
the
instantaneous population growth rate at low population
densities, and
both parameters are constrained to be positive. This is a
stochastic version
of the Ricker-logistic population growth model
\citep{ricker:1954}. We also
implemented a modified Gompertz-logistic density-dependent
model \citep{hart_gotelli:2011}:
\begin{equation}
N_{it} \sim
\text{Pois}(N_{it-1}\exp(r(1-\log(N_{it-1}+1)/\log(K+1))))
\label{eq:gomp}
\end{equation}
Here the interpretations of $r$ and $K$ are similar to in the
Ricker-logistic model.

Because a single Poisson distribution controls the
distribution of $N_{it}$ in each of these models, the discrete
convolution used by \citet{dail_madsen:2011} is not required,
speeding up processing time.

@@

The geometric-recruitment, exponential, Ricker-logistic, and Gompertz-logistic
versions of the DM models all share a common feature (or bug):
once the population at a site reaches 0, it must remain at 0.
This is because all contributions to population growth are
local in these models.  We generalized each these models that
include both internal and external (immigration) contributions
to population growth.  The exponential plus immigration
model is:
\begin{equation}
  N_{it} \sim \text{Poisson}(\exp(r)N_{it-1} + \iota)
  \label{eq:expimm2}
\end{equation}
where $\iota$ represents the immigration rate, and is
constrained to be
positive (this is equivalent to separate Poisson processes for
growth and immigration).  %This model is close to
%the constant-recruitment DM model (Eq.~\ref{eq:Nt}), with $\exp(r)$ instead
%of $\omega$ and $\iota$ instead of $\gamma$, except that the first
%process is Poisson distributed instead of binomial. 
The geometric-recruitment, Ricker-logistic,
and Gompertz-logistic models can be extended to allow for immigration in the
same way.   Deterministic versions of the Ricker-logistic + immigration and the 
Gompertz-logistic + immigration do reach stable equilibriums, but the equilibriums 
are not at $K$ and cannot be solved for analytically \citep{otto_day:2007}.  Therefore, we
refer to $K$ in these models as the semi-equilibrium abundance.
\begin{comment}
Similarly, the Ricker-logistic + immigration model can be represented as:
the
Gompertz-logistic + immigration model as: (13) and the geometric-recruitment +  immigration model as: (14)
\end{comment}

@@

\citet{dail_madsen:2011} suggested two distributions for
modeling
initial abundance: Poisson and negative binomial. We have
extended
their model to include another distribution, the zero-inflated
Poisson. This could be useful when, for example, one is modeling
the
abundance of several species with the same set of point
count surveys for all species, but this set includes routes
sites
outside the range of some of the species. The distribution
of
initial abundances can be represented as:
\begin{equation}
N_{i1} \sim \left\{
\begin{aligned}
0 &\; \text{with probability} \; \psi \\
\Lambda &\; \text{with probability} \; (1-\psi)
\end{aligned} \right.
\label{eq:ZIP}
\end{equation}
where $\psi$ represents the proportion of extra zeros.

>>>>

\author{Jeffrey A. Hostetler\\ %\footnote{Contact:
%     \url{HostetlerJ@si.edu}, +001-202-633-4205}\\
        \small Migratory Bird Center \\
        \small Smithsonian Conservation Biology Institute \\
        \small National Zoological Park, MRC 5503 \\
        \small Washington, DC 20013-7012\\
   \and Richard B. Chandler \footnote{Contact:
     \url{rbchan@uga.edu}, +001-706-542-5815} \\
        \small USGS Patuxent Wildlife Research Center \\
        \small Laurel, MD 20708}
\maketitle

\abstract{
\noindent
\begin{enumerate}
\item Models of population dynamics play a central role in theoretical and
applied ecology, where they are used for purposes such as testing 
hypotheses about density dependence and predicting species' responses 
to future environmental change or conservation actions. Failure to
account for observation error in such models can result in 
bias. Therefore, ecologists have increasingly relied on state-space
models to directly model both the observation error and the ecological
process of interest.
\item Conventional state-space models, however, have
three important limitations: (1) the parameters are not identifiable
in many common situations, (2) they do not admit spatial variation in
population dynamics, and (3) we have no clear interpretation of the
observation error. We demonstrate how each of these problems can be
resolved using a class of hierarchical models proposed by
\citet[Biometrics]{dail_madsen:2011} for spatially-replicated
time-series data that attributes observation error to imperfect detection.
\item We expand this class of models to accommodate classical growth
models (e.g. exponential and Ricker-logistic), zero-inflation,
and random effects such as observer-specific
detection probabilities. We also present methods for forecasting
population size under future environmental conditions. Implementation
of these ideas is possible using either frequentist or Bayesian
methods. We provide code to fit these models using the \textbf{R} package
\texttt{unmarked} and \textbf{JAGS} is also provided.
\item A simulation study indicated that bias was negligible and coverage
nominal for the proposed model extensions. An analysis of data from
the North American Breeding Bird Survey highlighted how these methods
can be readily applied to existing data, but it also suggested 
that precision would be low when we lack direct information about detection
probability (such as is collected using distance sampling or
replicated counts).
\end{enumerate}
}

@@

Theoretical ecology requires models of population dynamics for testing
hypotheses regarding spatial and temporal variation in
abundance. For example, much theoretical work has focused on
understanding the importance and existence of phenomenon such as
density-dependent population regulation, population cycling, and spatial synchrony
\citep{may:1975,turchin:1990,bjornstad_etal:1999}, %,royama:1977,dennis_taper:1994
and population models are required to evaluate associated hypotheses.
In applied contexts, population models are used for estimating extinction probabilities
\citep{schoener_spiller:1992,nadeem_lele:2011,hostetler_etal:2012} and for predicting the
effects of future environmental conditions or conservation actions on
population size \citep{jamieson_brooks:2004,hatfield_etal:2012}.
In order to address these questions, two complicating factors must
be confronted when fitting population models to data. 
First, deterministic models of population dynamics are virtually always inadequate due to process
variation, the inherent stochasticity in demographic parameters and environmental
conditions \citep{bjornstad_grenfell:2001,saether_engen:2002}.
Second, abundance---the natural state variable in studies
of population dynamics---can rarely be observed perfectly in field
studies because of observation error, such as imperfect
detection \citep{link_nichols:1994,kery_etal:2009}.

@@

A simple state-space models can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a population growth model, $\mu(N_{t-1})$
and random process variation denoted $\grave{o}_t$.
The growth model may be density-dependent, as in the case of the 
Ricker-logistic model, or it might be density-independent, such as when growth
is exponential.
%, in which $\mu(N_{t-1}) = N_{t-1}e^{r}$ where $r$ is the intrinsic rate of increase.
The full model can now be written as:
\begin{subequations}
  \label{eq:ss1}
  \begin{align}
    N_1 &= X_1 \label{eq:ss1a} \\
N_t &= \mu(N_{t-1}) + \grave{o}_{t-1} \quad \text{for} \;
t=2,\hdots,T \label{eq:ss1b} \\
X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \;
t=1,\hdots,T \label{eq:ss1c}
  \end{align}
\end{subequations}
where $\grave{o}_t$ is the random effect allowing for process
variation unaccounted for by the deterministic model. In classical
state-space models, the two sets of random effects
are assumed to be i.i.d. Gaussian deviates:
$\grave{o}_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. Standard practice is to
ignore process variation associated with $N_1$, as
indicated by Eq~\ref{eq:ss1a}. Alternatively, the population is
assumed to be at equilibrium such that the $N_1$ can be regarded as an
outcome of the equilibrium distribution.

Even though state-space models such as that shown in Eq.~\ref{eq:ss1}
are among the most widespread approaches for modeling population dynamics
using time-series data, several problems exist. First,
time-series data and the underlying abundance parameters of interest
are typically integer-valued, as in the case of count data, raising
concerns about the use of Gaussian distribution for the random
effects. Second, the model for observation error has little biological basis. 
Third, the model does not account for spatial variation in abundance. Lastly, some of the
parameters of the model are not estimable. We briefly discuss each of
these points before describing a general approach to resolving these
shortcomings.

Use of the Gaussian distribution for modeling random process variation
and observation error is motivated by convenience rather than
biology. Specifically, the Gaussian assumptions
allow for parameter estimation using the Kalman filter
\citep{dennis_etal:2006}, which is much more computationally efficient
than estimation methods when random effects are not Gaussian distributed
\citep{devalpine_hastings:2002}. %The problem with this is that it
This approach allows for negative and non-integer values of the state-variable, which is
inconsistent with the observed data, and may result in implausible predictions.
Transforming the count data and the underlying abundance variables do
little to resolve these issues \citep{ohara_kotze:2010}.

Another problem with standard state-space models is that the
observation error has no clear interpretation. The use of a mean zero
Gaussian distribution implies that $X_t$ will be
higher than $N_t$ as often as it is lower than $N_t$. 
Identifying a biological mechanism that would cause such symmetric errors
is difficult.
A more likely form of observation error, and one that has been recognized for well
over a century, results from failing to detect individuals that are
present. Imperfect detection may be attributable to
characteristics of the species under study, such as its elusiveness,
or to individuals being missed by the ecologist collecting the data in the field.
Although a vast number of methods have been devised for accounting for
this form of observation error, rarely have these methods been
integrated into state-space models \citep[but
see][]{buckland_etal:2004}.

Common formulations of state space models do not admit
spatial variation. This reduces the scope of the inferences that can
be drawn from the models. Moreover, space is an important factor in the
regulation of natural populations \citep{murdoch:1994,
pulliam:1988, gill_etal:2001}.

A more serious problem %than the ones associated with the Gaussian assumptions 
is that the parameters of the simple state-space models
such as Eq~\ref{eq:ss1} are not identifiable in many
circumstances \citep{polansky_etal:2009}. Recall that the
Eq.~\ref{eq:ss1} did not specify distributions for $N_1$, % or $X_1$;
which $N_1$ is a random variable. Thus, uncertainty exists
that should be accounted for in the model.  
%are needed to fully specify the model.
%To be more specific, a fully-specified state-space model requires at least
%three probability distributions, which we represent
%using bracket notation (i.e., $[Y|\Omega]$ is interpreted
%as the probability distribution of the random variable $Y$ given the
%parameter $\Omega$). The three probability distributions required for a
%state-space model are:
%%%% JEFF, does this extra sentence help? This notation comes from
%%%% several of the original state-space papers and is widespread in
%%%% books like Andy's and Bill's.
%\hl{[I'm not familiar with bracket notation,
% so these equations are confusing to me. How widely used is this
% notation in ecology? Will others likely understand this? Also, we might want
%to save space by trimming a bunch of this discussion back down.]}
%\begin{subequations}
%  \label{eq:ss2}
%  \begin{align}
%    [N_1&|\bm{\theta}] \label{eq:ss2a} \\ \label{eq:ss2b}
%[N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\
%\label{eq:ss2c}
%    [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
%  \end{align}
%\end{subequations}
%where $\bm{\theta}$ are the process variation parameters for the
%initial state, $\bm{\Theta}$ are the process variation parameters
%influencing how abundance changes over time, and $\bm{p}$ are the
%observation error parameters. As mentioned previously, Eq.~\ref{eq:ss2b}
%and Eq.~\ref{eq:ss2c} are assumed to be Gaussian in classical state-space
%models, but what should be the distribution for Eq.~\ref{eq:ss2a}?
%The three lines of general model
%described by Eq.~\ref{eq:ss2} correspond to the three equations shown
%for the specific example in Eq~\ref{eq:ss1}. However, in this case,
%we allow $N_1$ to be a random variable. But what should its
%istribution, $[N_1|\bm{\theta}]$ be?
%And how could $\bm{\theta}$ be estimated given that there is only a single observation
%available? Often, these problems are simply assumed away. For
%example, sometimes researchers assume that there is no process variation in
%the first year, although this seems unjustified. Alternatively, the
%population may be assumed to be at equilibrium such that the expected
%value of $N_t$ is constant through time. Although this 
%makes the parameters $\bm{\theta}$ identifiable, assuming equilibrium
%defeats the objective of many studies of population dynamics, namely
%determining why a population varies over time.

@@

The DM model is an extension of the $N$-mixture model
\citep{royle:2004biom}, which allows for inference about spatial
variation in abundance when individuals cannot be detected with
certainty. To estimate parameters associated with both abundance 
and the detection process, the original $N$-mixture model uses replicate
observations at each site. The observations are collected during sufficiently
short time intervals such that the population can safely be
assumed to be closed with respect to births, deaths, and movement. The DM
model relaxes this closure assumption and includes explicit parameters
describing population change over time.

@@

Like traditional state-space models, the DM model includes the
three conditionally related processes shown in Eq.~\ref{eq:ss1},
which correspond to (1) initial abundance, i.e. the
abundance at site $i$ during the first primary period,
denoted (2) abundance at time $t$ (for $t>1$) which depends upon
abundance at $t-1$, and (3) the
detection process \citep{dail_madsen:2011}.
The first two processes describe the state process---the
variation in abundance in space and time. The third process %simply 
describes the the relationship between
abundance and the observed count data.

@@

The DM model assumes that abundance in time $t$ is a function of
abundance in time $t-1$, i.e. abundance at each site evolves as a
first order Markovian process. %, although higher order processes are also possible.
\citet{dail_madsen:2011} considered several models to describe the temporal dynamics;
however, in each case they modeled $N_t$ as the sum of two random variables:
$S_{it}$, the number of indivduals surviving from $t-1$ and not
emigrating; and $G_{it}$ the number of new individuals entering
the population. Their most general model was
\begin{equation}
\left.\begin{aligned}
S_{it}|N_{it-1} &\sim \mathrm{Bin}(N_{it-1}, \omega) \\
G_{it}|N_{it-1} &\sim \mathrm{Pois}(\gamma(N_{it-1})) \\
N_{it} &= S_{it}+G_{it}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
where $\omega$ is the apparent survival probability and $\gamma$
is the recruitment rate (which can depend on $N_{it-1}$).
\citet{dail_madsen:2011} proposed three
models for $\gamma$: the constant model,
$G_{it} \sim \mathrm{Pois}(\gamma)$ where recruitment does not
depend on $N_{it-1}$, and which simulates a ``propagule rain'' of new
individuals; the autoregressive model, $G_{it} \sim
\mathrm{Pois}(\gamma(N_{it-1}))$, which
simulates geometric or density independent growth; and the
``no-trend'' model, $\gamma = (1-\omega)\Lambda$, which keeps
expected abundance constant over time. Covariates of
$\omega$ and $\gamma$ can be easily accommodated, for example
using logit- and log-linear models respectively.

@@

Bayesian inference, however, has several appealing features. First, it
allows direct probability statements to be made about a hypothesis
given data \citep{link_barker:2010}.    
Second, Bayesian methods offer straight-forward approaches for
combine data from multiple sources or use existing estimates of parameters as prior
distributions. In practice, simulation methods such as Markov
chain Monte Carlo (MCMC) are used to simulate the posterior
distributions. %Although MCMC can be slow, our experience is that it
%may actually converge faster than the amount of time required to
%maximize the likelihood shown in Eq.\ref{eq:like}, especially
%when the upper bound of the summation must be high, e.g. $N_{max}>200$.
%Furthermore, 
MCMC may be the only possible option for estimating parameters
using some of the extensions described below, such as allowing for
additional random effects.

@@

%For example here is \textbf{BUGS} code for the ``geometric'' model described
%in Eq.~\ref{eq:hm}:
%\begin{verbatim}
%model {
%lambda ~ dunif(0, 5)
%omega ~ dunif(0,1)
%gamma ~ dunif(0, 10)
%p ~ dunif(0,1)
%for(i in 1:nSites) {
%  N[i,1] ~ dpois(lambda)
%  y[i,1] ~ dbin(p, N[i,1])
%  for(t in 2:nYears) {
%    S[i,t-1] ~ dbin(omega, N[i,t-1])
%    G[i,t-1] ~ dpois(gamma*N[i,t-1])
%    N[i,t] <- S[i,t-1] + G[i,t-1]
%    y[i,t] ~ dbin(p, N[i,t])
%    }
%  }
%}
%\end{verbatim}

@@

Partitioning population growth into survival and recruitment
is ideal in terms of providing a mechanistic description of population
dynamics, but this is not always possible using simple count data,
especially when the sites are not closed with respect to immigration
and emigration. %Furthermore, a
%simpler model would have several merits: faster running time, fewer
%total model combinations, and possibly more realistic estimates.
If the mechanistic model must be sacrificed, we suggest replacing it
with classical population growth models that have a long history in
ecology. The exponential growth model is the simplest: 
$f(N_{it-1}) = N_{it-1}e^r$ where $r$ is the intrinsic
rate of increase. %Although demographic and environmental stochasticity
%in $r$ could be modeled as functions of covariates or random effects, a
%simpler way of 
To allow for demographic stochasticity, we can regard $N_{it}$ as a Poisson 
random variable, simplifying Eq.~\ref{eq:Nt} to:
\begin{equation}
  N_{it} \sim \text{Pois}(\exp(r)N_{it-1}).
\label{eq:exp}
\end{equation}
This,
like the geometric-recruitment version of the model, is a variant on a simple
density-independent, exponential model of population growth.
Density-dependent versions of the model are also possible.  For
example:
\begin{equation}
  N_{it} \sim \text{Pois}(N_{it-1}\exp(r(1-N_{it-1}/K)))
\label{eq:rick}
\end{equation}
where $K$ is the stable equilibrium of the population and $r$ is
the
instantaneous population growth rate at low population
densities, and
both parameters are constrained to be positive. This is a
stochastic version
of the Ricker-logistic population growth model
\citep{ricker:1954}. We also
implemented a modified Gompertz-logistic density-dependent
model \citep{hart_gotelli:2011}:
\begin{equation}
N_{it} \sim
\text{Pois}(N_{it-1}\exp(r(1-\log(N_{it-1}+1)/\log(K+1))))
\label{eq:gomp}
\end{equation}
Here the interpretations of $r$ and $K$ are similar to in the
Ricker-logistic model.
Because a single Poisson distribution controls the
distribution of $N_{it}$ in each of these models, the discrete
convolution used by \citet{dail_madsen:2011} is not required,
speeding up processing time.

@@

The geometric-recruitment, exponential, Ricker-logistic, and Gompertz-logistic
versions of the DM models all share a common feature (or bug):
once the population at a site reaches 0, it must remain at 0
because all contributions to population growth are assumed to be
local.  We therefore generalized these models to
include both internal and external (immigration) contributions
to population growth.  The exponential plus immigration
model is:
\begin{equation}
  N_{it} \sim \text{Poisson}(\exp(r)N_{it-1} + \iota)
  \label{eq:expimm2}
\end{equation}
where $\iota$ represents the immigration rate, and is
constrained to be
positive (this is equivalent to separate Poisson processes for
growth and immigration).  %This model is close to
%the constant-recruitment DM model (Eq.~\ref{eq:Nt}), with $\exp(r)$ instead
%of $\omega$ and $\iota$ instead of $\gamma$, except that the first
%process is Poisson distributed instead of binomial. 
The geometric-recruitment, Ricker-logistic,
and Gompertz-logistic models can be extended to allow for immigration in the
same way.   Deterministic versions of the Ricker-logistic + immigration and the 
Gompertz-logistic + immigration do reach stable equilibriums, but the equilibriums 
are not at $K$ and cannot be solved for analytically \citep{otto_day:2007}.  Therefore, we
refer to $K$ in these models as the semi-equilibrium abundance.
\begin{comment}
Similarly, the Ricker-logistic + immigration model can be represented as:
the
Gompertz-logistic + immigration model as: (13) and the geometric-recruitment +  immigration model as: (14)
\end{comment}

@@

\citet{dail_madsen:2011} suggested two distributions for modeling
initial abundance: Poisson and negative binomial. We have
extended their model to include another distribution, the zero-inflated
Poisson. This could be useful when, for example, one is modeling
the abundance of several species with the same set of point
count surveys for all species, but some survey sites are
outside the range of some of the species. The distribution
of initial abundances can be represented as:
\begin{equation}
N_{i1} \sim \left\{
\begin{aligned}
0 &\; \text{with probability} \; \psi \\
\Lambda &\; \text{with probability} \; (1-\psi)
\end{aligned} \right.
\label{eq:ZIP}
\end{equation}
where $\psi$ represents the proportion of extra zeros.
